Estimating Access to Rabies Post-Exposure Prophylaxis in Madagascar

### 2019-03-14 accessPEP.Rmd
model comparison (corrected bug in district model)
+ sensitivity analyses etc.

## Files
source("R/utils.R")
source("R/models.R")
source("R/data_functions.R")
select <- dplyr::select
knitr::opts_chunk$set(fig.pos = 'h', echo = FALSE, message = FALSE, warning = FALSE, 
                      results = "hide",
                      include = TRUE)
```

## INTRODUCTION
Inequities in access to care are a major driver of disease burden globally. Often, the populations at greatest risk of a given disease are the most underserved. Delivering interventions to these groups is challenging due to financial and infrastructural limitations, and requires careful considerations of how best to optimize the allocation of limited resources.

Canine mediated rabies is estimated to cause approximately 60,000 human deaths anually. These deaths can be prevented through prompt administration of post-exposure prophylaxis. However, access to this intervention is highly limited in areas where the disease is endemic (*cite GAVI paper or Nandini's paper of geographic availability of PEP*). GAVI bit here...potential to greatly reduce disease burden.

Data on true rabies exposures in humans and incidence in animals is also lacking in most of these countries, with the most commonly available data being numbers of bite victims reporting to health facilities. The majority of rabies burden studies thus far have integrated these data on reported bites into a probability decision tree framework to estimate rabies burden, often with the key assumption that overall reported bite incidence (i.e. both bites due to non-rabid and rabid animals) are proportional to rabies incidence (i.e. the more bites reported in a location, the higher the incidence of *rabies* exposures there) and that reporting is uniform across space. While at the national level these estimates may be accurate, at the sub-national level, this framework will likely underestimate rabies deaths in places with low reporting and overestimate rabies deaths in places with high reporting of bites.

In Madagascar, Institut Pasteur de Madagascar (IPM) provides PEP at no-cost to patients at 31 clinics across the country. PEP is not available at any other public clinics or through the private sector. In addition, there is limited control in dog populations and the disease is endemic throughout the country. Due to the spatially restricted nature of PEP, geographic access is likely to be a major driver of disease burden within the country. To get spatial estimates of disease burden in this context, we flip the standard decision tree and make the assumption that reported bite incidence reflects access to PEP rather than differences in rabies incidence, using travel times to clinics as a predictor of reported bites. Then using a range of rabies incidence given endemic transmission with no mass dog vaccination (*GAVI paper*), we generate sub-national estimates of rabies burden in an adapted decision tree framework. Finally, using this same model pipeline, we explore the impacts of geographically expanding access to PEP in Madagascar on reducing the burden of human rabies deaths.

@@

### Bite patient data
We used a database of bite patient forms submitted to IPM from ARMC across the country between 2014 - 2017. These were individual patient data forms that were submitted as frequently as monthly to annually by clinics. Of the 31 existing ARMC, `r length(no_data) - 2` submitted fewer than 10 forms over the four years. Two clinics, the IPM ARMC and the Fianarantsoa ARMC had separate databases which were not available at the time of analysis. Overall, we had data from `r 31 - length(no_data)` clinics across the country (Fig 1A). These data include details of the administrative district of the bite patient's address and the date of reporting. We also had `r max(ctar$month, na.rm = TRUE)` months of data from the Moramanga District that were resolved to the commune level (the administrative level below the district).

@@

Previous work in the Moramanga District showed that low risk contacts with probable or confirmed rabies cases make up approximately 20% of patients reporting to ARMC (Rajeev et al. 2018). Generally, contacts present as clustered cases, so we excluded patients reporting on any dates that had greater than 3 standard deviations above the mean number of patients reporting per day (Figure 3A). We validated this method using the Moramanga ARMC data for which we had details on the type of exposure, and found that setting the threshold to 3 standard deviations (SDs) resulted in approximately 50% of known contacts excluded, with only 2% of non-contacts excluded (Figure S4). For the national data for which a subset of patient forms were explicitly noted to be contacts, we found that our exclusion criteria of 3 SDs identified `r round((ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 & exclude_bydate == 1]))/length(ctar_check %$% exclude_bycomment[exclude_bycomment == 1])*100, 2)` of known contacts. We further excluded these known contacts which were not identified based on the daily distribution of patients, resulting in the exclusion of approximately `r round((ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 | exclude_bydate == 1]))/nrow(ctar_check)*100, 2)`% of patient records from the national data. We also compared this method to assuming that 20% of total bites at the district level were contacts.

@@

After excluding contacts and correcting for undersubmission of forms, our final dataset consisted of estimates of average bite incidence for `r nrow(exps_dist)` districts from `r length(unique(exps_dist$dist_catch))` catchments (Figure 4).

@@

```{r models}
model.bites <- function(bites, names_bites, covar, pop, names_covar, sum = TRUE, 
                        pop_predict = "addPop", covar_name = "ttimes_weighted",
                        beta, beta_pop, intercept, trans, run = "optim",...)

## Exclude for comm data!
comm_covars <- filter(comm_data, exclude == 0)

pop_predict <- c("addPop", "onlyPop", "none")
pop_vars <- list(pop_dist = exps_dist$pop, pop_commune = comm_covars$pop)
covars <- list(distance_district = exps_dist$distance, distance_commune = comm_covars$distance, 
               ttimes_district = exps_dist$ttimes_weighted_dist/60, ttimes_commune = comm_covars$ttimes_weighted/60)
sum_it <- c(FALSE, TRUE, FALSE, TRUE)
run_type <- "optim"

mods_mada <- 
  foreach(i = 1:length(pop_predict), .combine = c) %:%
  foreach(j = 1:length(sum_it), .combine = c) %:%
  foreach(k = 1:length(covars), .combine = c) %do% {
  i = 1; j = 1; k = 1;

  if(sum_it[j] == TRUE) {
    names_covar <- comm_covars$mdg_dis_co
  } else {
    names_covar = ""
  }
  data_est <- list(bites = exps_dist$bites, run = run_type, covar = covars[[k]],
                   names_bites = exps_dist$district,
                   covar_name = names(covars)[k], names_covar = names_covar,
                   trans = 1e5)
  start_params <- list(beta = 1e-5, intercept = 0.1)

  if(pop_predict[i] == "addPop") {
    ## need beta_pop
    start_params <- c(start_params, beta_pop = 1e-5)
  }

  if(grepl("commune", names(covars)[k], fixed = TRUE)) {
    data_est <- c(data_est, list(pop = pop_vars$pop_commune))
  } else {
    data_est <- c(data_est, list(pop = pop_vars$pop_dist))
  }

  check <- mle2(model.bites, start = start_params, data = data_est)

  attributes(check)$name <- paste(scale[i], sum_name, covar_name)
  check
}

mods_moramanga
mods_pop_mada_moramanga

## all mods = combine

prof_CI <- lapply(mods, confint)

mada_mod_predicts <- as.data.frame(list(beta = mapply(function(x) coef(x)["beta"], mods), 
     intercept = mapply(function(x) coef(x)["intercept"], mods), 
     beta_upper = mapply(function(x) x["beta", 2], prof_CI),
     beta_lower = mapply(function(x) x["beta", 1], prof_CI),
     intercept_upper = mapply(function(x) x["intercept", 2], prof_CI),
     intercept_lower = mapply(function(x) x["intercept", 1], prof_CI),
     likelihood = mapply(function(x) logLik(x), mods),
     model = mapply(function(x) attributes(x)$name, mods)))

mada_mod_results <- as.data.frame(list(params = c("beta", "intercept"), 
                   estimates = t(mapply(function(x) coef(x), mods)), 
                   upper = t(mapply(function(x) x[, 2], prof_CI)),
                   lower = t(mapply(function(x) x[, 1], prof_CI)), 
                   data = "Mada"))

## Predictions
library(foreach)

mods <- 
  foreach(l = 1:nrow(mada_), .combine = c) %do% {
      if(pop_predict[k] == TRUE) covar_name = "pop" else covar_name = names(covar_mada)[l]
      if (scale[i] == "Mada") {
        if(sum[j] == TRUE) {
          sum_name <- "summed"
          check <- predict.bites(bites = exps_dist$bites, 
                           names_bites = exps_dist$district, 
                           pop = exps_dist$pop,
                           pop_sub = comm_data$pop, 
                           names_pop = comm_data$mdg_dis_co, 
                           names_covar = comm_data$mdg_dis_co, trans = 1e5,
                           covar_sub = covar_mada_sub[[l]],
                           sum = sum[j], pop_predict = pop_predict[k])
        } else {
          sum_name <- ""
          check <- mle2(get.likelihood.bites, start = list(beta = 1e-5, intercept = 0.1),
               data = c(data_mada, list(covar = covar_mada[[l]]),
                           sum = sum[j], pop_predict = pop_predict[k]))
        }
      }   
      if(scale[i] == "Mora") {
        if(sum[j] == TRUE) {
          check <- NULL
        } else {
          sum_name <- ""
          check <- mle2(get.likelihood.bites, start = list(beta = 1e-5, intercept = 0.1),
             data = c(data_mora, list(covar = covar_mora[[l]]),
                           sum = sum[j], pop_predict = pop_predict[k]))
        }
      }
  attributes(check)$name <- paste(scale[i], sum_name, covar_name)
  check
}

ttimes_plot <- seq(0, 720, by = 30)
pop_plot <- seq(100, 1e6, length.out = length(ttimes_plot))
expect <- foreach(i = 1:nrow(mada_mod_predicts), .combine = rbind) %do% {
  check <- mada_mod_predicts$expect_fun[[i]]
  exp <- check(beta = mada_mod_predicts$beta[i], intercept = mada_mod_predicts$intercept[i],
               covar = ttimes_plot/60, pop = 1e5, pop_covar = pop_plot, 
               trans = 1e5)
  exp_upper <- check(beta = mada_mod_predicts$beta_upper[i], 
                     intercept = mada_mod_predicts$intercept_upper[i],
                     covar = ttimes_plot/60, pop = 1e5, pop_covar = pop_plot, trans = 1e5)
  exp_lower <- check(beta = mada_mod_predicts$beta_lower[i], 
                     intercept = mada_mod_predicts$intercept_lower[i],
                     covar = ttimes_plot/60, pop = 1e5, pop_covar = pop_plot, trans = 1e5)
  as.data.frame(list(ttimes = ttimes_plot, 
                     pop = pop_plot, exp = exp, exp_upper = exp_upper, exp_lower = exp_lower, 
                     model = rownames(mada_mod_predicts)[i], data = mada_mod_predicts$data[i]))
}

ggplot(data = preds, aes(x = observed, y = predicted, color = model)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = 2, color = "grey")
ggplot(data = filter(expect, !grepl("pop", model, fixed = TRUE)), 
       aes(x = ttimes, y = exp, color = model)) +
  geom_line() +
  geom_ribbon(aes(ymin = exp_lower, ymax = exp_upper, fill = model), alpha = 0.5)

AICctab(mada_ttimes_fit, mada_distance_fit, mada_pop_fit, mada_sumttimes_fit, mada_sumpop_fit,
        mada_sumdist_fit, nobs = nrow(exps_dist))

par_df <- data.frame(model = c("Mada communes", "Moramanga communes", "Mada districts", 
                               "Mada communes", "Moramanga communes", "Mada districts"),
                     par = c("B_ttimes", "B_ttimes", "B_ttimes", "intercept", "intercept",
                             "intercept"),
                     values = c(B_ttimes_mada, B_ttimes_mora, B_ttimes_dist, 
                                B_0_mada, B_0_mora, B_0_dist),
                     upper_CI = c(CI = B_ttimes_mada_CI[2], B_ttimes_mora_CI[2],
                                  B_ttimes_dist_CI[2],
                                  B_0_mada_CI[2], B_0_mora_CI[2], B_0_dist_CI[2]),
                     lower_CI = c(CI = B_ttimes_mada_CI[1], B_ttimes_mora_CI[1],
                                  B_ttimes_dist_CI[1], 
                                  B_0_mada_CI[1], B_0_mora_CI[1], B_0_dist_CI[1]))
kable(par_df, caption = "Table 1: summary of parameter estimates")
ttimes_plot <- seq(0, 15*60, by = 30)
pop_plot <- 1e5

## Mada
par_df <- as.data.frame(list(model = c("Mada communes", "Moramanga communes", "Mada districts"),
                     B_ttimes = c(B_ttimes_mada, B_ttimes_mora, B_ttimes_dist),
                     B_0 = c(B_0_mada, B_0_mora, B_0_dist),
                     B_0_upper = c(B_0_mada_CI[2], B_0_mora_CI[2], B_0_dist_CI[2]),
                     B_0_lower = c(B_0_mada_CI[1], B_0_mora_CI[1], B_0_dist_CI[1]),
                     B_ttimes_upper = c(CI = B_ttimes_mada_CI[2], B_ttimes_mora_CI[2],
                                  B_ttimes_dist_CI[2]),
                     B_ttimes_lower = c(CI = B_ttimes_mada_CI[1], B_ttimes_mora_CI[1],
                                  B_ttimes_dist_CI[1])))
```

```{r model table, results = "show"}
kable(par_df, caption = "Table 1: summary of parameter estimates")
```

```{r figure 5B, fig.cap="Figure 5B"}
## Data vs. predicted
comm_data %>%
      filter(exclude == 0) %>%
      mutate(bites = exp(B_ttimes_mada*ttimes_weighted + B_0_mada)*pop) %>%
      group_by(mdg_dis_co) %>%
      summarize(sum_bites = sum(bites)) %>%
      left_join(select(exps_dist, district, dist_bites = bites), 
                by = c("mdg_dis_co" = "district")) %>%
      filter(!is.na(dist_bites)) -> sum_bites
sum_bites %$% plot(dist_bites, sum_bites, bty = "n", pch = 20, col = alpha("darkblue", 0.5),
                   xlab = "Observed mean bites", ylab = "Model predicted mean bites", 
                   xlim = c(0, 5000))
abline(a = 0, b = 1, col = "grey", lty = 2)

exps_dist %>%
      mutate(bites_mu = exp(B_ttimes_dist*ttimes_weighted_dist + B_0_dist)*pop) -> check
points(check$bites, check$bites_mu, col = alpha("purple", 0.5), pch = 20)

mora_bites %>%
      mutate(bites_mu = exp(B_ttimes_mora*ttimes_weighted + B_0_mora)*pop) -> check
points(check$bites, check$bites_mu, bty = "n", pch = 20, col = alpha("red", 0.5))
legend("topright", inset = c(-0.1, -0.2),
       c("Mada communes", "Mada districts", "Moramanga communes"), 
       col = c(alpha("darkblue", 0.5), alpha("purple", 0.5), alpha("red", 0.5)), 
       pch = 20, 
       bty = "n", xpd = TRUE)

@@

#### Model validation
```{r model validation}
## Models
get.likelihood.other <- function(weight, bites, pop, params) {
  B_ttimes <- params[1]
  B_0 <- params[2]
  exp_bites <- inv.logit(B_ttimes*weight + B_0)*pop
  return(-sum(dpois(round(exp_bites), lambda = round(bites), log = TRUE)))
}

## Distance to CTAR
ctar_coords <- cbind(ctar_metadata$LONGITUDE, ctar_metadata$LATITUDE)
district_coords <- coordinates(mada_district)
distance_mat <- distm(district_coords, ctar_coords)
dist_min <- apply(distance_mat, 1, min)/1000
exps_dist$dist_min <- dist_min[match(exps_dist$district, mada_district$mdg_dis_co)]

## Unweighted travel times
exps_dist$ttimes_unweighted <- mada_district$study_area[match(exps_dist$district,
                                                              mada_district$mdg_dis_co)]

### District parameters
dist_pars <- optim(par = c(1e-5, -0.1), get.likelihood.other, 
                   weight = exps_dist$dist_min, bites = exps_dist$bites, pop = exps_dist$pop,
                   hessian = TRUE)
B_ttimes_dist <- dist_pars$par[1]
B_0_dist <- dist_pars$par[2]

## Getting standard errors
# this is how you can get standard errors from optim, but inverting the hessian
se <- sqrt(abs(diag(solve(dist_pars$hessian)))); se

#simulate "standard errors"
varcov <- solve(dist_pars$hessian)

ntest = 1000; testpar = dist_pars$par; sim.par = matrix(NA, ntest, 2)
for (k in 1:ntest) {
  sim.par[k, ] = mvrnorm(1, dist_pars$par, varcov)
}

B_ttimes_dist_CI <- quantile(sim.par[,1],c(0.025,0.975))
B_0_dist_CI <- quantile(sim.par[,2],c(0.025,0.975))

plot(exps_dist$bites, inv.logit(B_ttimes_dist*exps_dist$dist_min+ B_0_dist)*exps_dist$pop)
plot(exps_dist$ttimes_weighted_dist, exps_dist$bites)

### Unweighted
dist_pars <- optim(par = c(1e-3, 0.1), get.likelihood.other, 
                   weight = exps_dist$ttimes_unweighted, bites = exps_dist$bites, 
                   pop = exps_dist$pop,
                   hessian = TRUE)
B_ttimes_dist <- dist_pars$par[1]
B_0_dist <- dist_pars$par[2]

## Getting standard errors
# this is how you can get standard errors from optim, but inverting the hessian
se <- sqrt(abs(diag(solve(dist_pars$hessian)))); se

#simulate "standard errors"
varcov <- solve(dist_pars$hessian)

ntest = 1000; testpar = dist_pars$par; sim.par = matrix(NA, ntest, 2)
for (k in 1:ntest) {
  sim.par[k, ] = mvrnorm(1, dist_pars$par, varcov)
}

B_ttimes_dist_CI <- quantile(sim.par[,1],c(0.025,0.975))
B_0_dist_CI <- quantile(sim.par[,2],c(0.025,0.975))

plot(exps_dist$bites, exp(B_ttimes_dist*exps_dist$ttimes_unweighted+ B_0_dist))
plot(exps_dist$ttimes_weighted_dist, exps_dist$dist_min)

### Pop
get.likelihood.pop <- function(bites, pop, params) {
  B_ttimes <- params[1]
  B_0 <- params[2]
  exp_bites <- exp(B_ttimes*pop/1e5 + B_0)
  return(-sum(dpois(round(exp_bites), lambda = round(bites), log = TRUE)))
}
pop_pars <- optim(par = c(1e-5, 0.1), get.likelihood.pop, 
                   pop = exps_dist$pop, bites = exps_dist$bites, 
                   hessian = TRUE)
B_pop <- pop_pars$par[1]
B_0_pop <- pop_pars$par[2]
check <- glm(round(bites) ~ pop, data = exps_dist, family = "poisson")
plot(exps_dist$bites, exp(B_pop*exps_dist$pop/1e5 + B_0_pop), pch = 20, col = alpha("red", 0.5))
plot(exps_dist$bites, predict(check, type = "response"), pch = 20, col = alpha("orange", 0.5))
abline(a = 0, b = 1, col = "grey", lty = 2)

## Getting standard errors
# this is how you can get standard errors from optim, but inverting the hessian
se <- sqrt(abs(diag(solve(pop_pars$hessian)))); se

#simulate "standard errors"
varcov <- solve(pop_pars$hessian)

ntest = 1000; testpar = pop_pars$par; sim.par = matrix(NA, ntest, 2)
for (k in 1:ntest) {
  sim.par[k, ] = mvrnorm(1, pop_pars$par, varcov)
}

B_pop_CI <- quantile(sim.par[,1],c(0.025,0.975))
B_0_pop_CI <- quantile(sim.par[,2],c(0.025,0.975))

## @ commune level
get.likelihood.mada <- function(commune, district, params) {
  # To test likelihood function
  # commune = comm_data
  # district = exps_dist
  # params = c(1e-6, 0.1)
  B_ttimes <- params[1]
  B_0 <- params[2]
  commune %>%
      filter(exclude == 0) %>%
      group_by(mdg_dis_co) %>%
      mutate(bites = exp(B_ttimes*pop/1e5 + B_0)) %>%
      summarize(sum_bites = sum(bites)) %>%
      left_join(select(district, district, dist_bites = bites), 
                by = c("mdg_dis_co" = "district")) %>%
      filter(!is.na(dist_bites)) -> sum_bites
  return(-sum(dpois(round(sum_bites$dist_bites), lambda = sum_bites$sum_bites, log = TRUE)))
}
comm_pars <- optim(par = c(1e-3, 0.1), get.likelihood.mada, 
                   commune = comm_data, district = exps_dist, 
                   hessian = TRUE)
B_ttimes_comm <- comm_pars$par[1]
B_0_comm <- comm_pars$par[2]

## Getting standard errors
# this is how you can get standard errors from optim, but inverting the hessian
se <- sqrt(abs(diag(solve(comm_pars$hessian)))); se

#simulate "standard errors"
varcov <- solve(comm_pars$hessian)

ntest = 1000; testpar = comm_pars$par; sim.par = matrix(NA, ntest, 2)
for (k in 1:ntest) {
  sim.par[k, ] = mvrnorm(1, comm_pars$par, varcov)
}

B_ttimes_comm_CI <- quantile(sim.par[,1],c(0.025,0.975))
B_0_comm_CI <- quantile(sim.par[,2],c(0.025,0.975))

comm_data %>%
      filter(exclude == 0) %>%
      mutate(bites = exp(B_ttimes_comm*pop/1e5 + B_0_comm)) %>%
      group_by(mdg_dis_co) %>%
      summarize(sum_bites = sum(bites)) %>%
      left_join(select(exps_dist, district, dist_bites = bites), 
                by = c("mdg_dis_co" = "district")) %>%
      filter(!is.na(dist_bites)) -> sum_bites
sum_bites %$% plot(dist_bites, sum_bites, bty = "n", pch = 20, col = alpha("darkblue", 0.5),
                   xlab = "Observed mean bites", ylab = "Model predicted mean bites", 
                   xlim = c(0, 5000))
comm_data %>%
      filter(exclude == 0) %>%
      mutate(bites = exp(B_ttimes_mada*ttimes_weighted + B_0_mada)*pop) %>%
      group_by(mdg_dis_co) %>%
      summarize(sum_bites = sum(bites)) %>%
      left_join(select(exps_dist, district, dist_bites = bites), 
                by = c("mdg_dis_co" = "district")) %>%
      filter(!is.na(dist_bites)) -> sum_bites
sum_bites %$% plot(dist_bites, sum_bites, pch = 20, col = alpha("red", 0.5))
abline(a = 0, b = 1, col = "grey", lty = 2)

exps_dist %>%
      mutate(bites_mu = exp(B_ttimes_dist*ttimes_weighted_dist + B_0_dist)*pop) -> check1
points(check1$bites, check1$bites_mu, col = alpha("purple", 0.5), pch = 20)
points(exps_dist$bites, predict(check, type = "response"), pch = 20, col = alpha("orange", 0.5))
abline(a = 0, b = 1, col = "grey", lty = 2)

mora_bites %>%
      mutate(bites_mu = exp(B_ttimes_mora*ttimes_weighted + B_0_mora)*pop) -> check
points(check$bites, check$bites_mu, bty = "n", pch = 20, col = alpha("blue", 0.5))
check <- glm(round(bites) ~ pop, data = mora_bites, family = "poisson")
plot(mora_bites$bites, predict(check, type = "response"), pch = 20, col = alpha("orange", 0.5))
abline(a = 0, b = 1, col = "grey", lty = 2)

## Correlation plot

## Table of fits/parameter results

## Impact of contacts

## Impact of RLE threshold

@@

```{r reporting by incidence, fig.cap="Figure 6"}
## Mada
bites_100k <- exp(B_ttimes_mada*ttimes_plot + B_0_mada)*pop_plot
R_100k_min <- pop_plot/25*0.01*0.39
R_100k_max <- pop_plot/5*0.01*0.39

plot(ttimes_plot/60, 
     bites_100k*(ifelse(bites_100k*0.6 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.6))/R_100k_min,
     type = "l", bty = "n",
     ylab = "Reporting of rabid exposures", xlab = "Travel times (hrs)", col = "darkred")
lines(ttimes_plot/60, 
     bites_100k*(ifelse(bites_100k*0.2 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.2))/R_100k_min,
     col = "darkred")
lines(ttimes_plot/60, 
     bites_100k*(ifelse(bites_100k*0.6 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.6))/R_100k_max,
     col = "red", lty = 2)
lines(ttimes_plot/60, 
     bites_100k*(ifelse(bites_100k*0.2 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.2))/R_100k_max,
     col = "red", lty = 2)
polygon(x = c(ttimes_plot/60, rev(ttimes_plot/60)), 
        y = c(bites_100k*(ifelse(bites_100k*0.2 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.2))/R_100k_min,
             rev(bites_100k*(ifelse(bites_100k*0.2 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.2))/R_100k_max)),
        border = NA, col = alpha("red", 0.2))
polygon(x = c(ttimes_plot/60, rev(ttimes_plot/60)), 
        y = c(bites_100k*(ifelse(bites_100k*0.6 > R_100k_min, 0.98*R_100k_min/bites_100k, 0.6))/R_100k_min,
             rev(bites_100k*(ifelse(bites_100k*0.6 > R_100k_max, 0.98*R_100k_max/bites_100k, 0.6))/R_100k_max)),
        border = NA, col = alpha("red", 0.6))
legend("topright", inset = c(-0.10, 0), c("Max rabies exposures", "Min rabies exposures"), lty = c(2, 1), 
       col = c("red", "darkred"),
       bty = "n", xpd = TRUE)
legend("topright", inset = c(0.01, 0.2), c("p_rabid = 0.2", "p_rabid = 0.6"), border = NA, 
       fill = c(alpha("red", 0.2), alpha("red", 0.6)), bty = "n", xpd = TRUE)
abline(v = max(ttimes_plot[bites_100k*0.6 > R_100k_min])/60, col = "grey", lty = 3)
abline(v = max(ttimes_plot[bites_100k*0.2 > R_100k_max])/60, col = "grey", lty = 3)
```

```{r scale example}
## Comparing scales
comm_data$reporting_min <- get.reporting_burden(pop = comm_data$pop, ttimes = comm_data$ttimes_weighted, 
                                            B_ttimes = B_ttimes_mada, B_0 = B_0_mada,
                                            hdr = 5, p_rabid = 0.2)

comm_data$reporting_max <- get.reporting_burden(pop = comm_data$pop, ttimes = comm_data$ttimes_weighted, 
                                            B_ttimes = B_ttimes_mada, B_0 = B_0_mada,
                                            hdr = 25, p_rabid = 0.6)
comm_data %>%
  group_by(district) %>%
  mutate(pop_total = sum(pop, na.rm = TRUE), prop_pop = pop/pop_total) %>%
  summarize(reporting_min = sum(reporting_min*prop_pop), 
            reporting_max = sum(reporting_max*prop_pop)) %>%
  mutate(reporting_med = (reporting_min + reporting_max)/2,
         type = "summed from commune") -> summed_rep_dist

mada_district@data %>%
  mutate(reporting_min = get.reporting_burden(pop = pop2015un, ttimes = ttimes_weighted, 
                                              B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 5, 
                                              p_rabid = 0.2), 
         reporting_max = get.reporting_burden(pop = pop2015un, ttimes = ttimes_weighted, 
                                              B_ttimes = B_ttimes_dist, B_0 = B_0_dist, hdr = 25, 
                                              p_rabid = 0.6),
         reporting_med = (reporting_min + reporting_max)/2,
         type = "district level") %>%
  select(district, reporting_min, reporting_max, reporting_med, type) -> rep_comp

rep_comp <- rbind(rep_comp, summed_rep_dist)
rep_comp$ttime <- mada_district$ttimes_weighted[match(rep_comp$district, mada_district$district)]

>>>>

## Files
source("R/utils.R")
source("R/models.R")
source("R/data_functions.R")
source("R/get.burden.R")
select <- dplyr::select
knitr::opts_chunk$set(fig.pos = 'h', echo = FALSE, message = FALSE, warning = FALSE, 
                      results = "hide",
                      include = TRUE)
```

## INTRODUCTION
Inequities in access to care are a major driver of disease burden globally. Often, the populations at greatest risk of a given disease are the most underserved. Delivering interventions to these groups is challenging due to financial and infrastructural limitations, and requires careful consideration of how best to optimize the allocation of limited resources.

Canine mediated rabies is estimated to cause approximately 60,000 human deaths anually. These deaths can be prevented through prompt administration of post-exposure prophylaxis. However, access to this intervention is highly limited in areas where the disease is endemic (*cite GAVI paper or Nandini's paper of geographic availability of PEP*). Data on true rabies exposures in humans and incidence in animals is also lacking in most of these countries, with the most commonly available data being numbers of bite victims reporting to health facilities. The majority of rabies burden studies thus far have integrated these data on reported bites into a probability decision tree framework to estimate rabies burden, often with the key assumption that overall reported bite incidence (i.e. both bites due to non-rabid and rabid animals) are proportional to rabies incidence (i.e. the more bites reported in a location, the higher the incidence of *rabies* exposures there) and that reporting is uniform across space. At the sub-national level where access to PEP is highly heterogenous, these assumptions likely underestimate rabies deaths in places with low reporting and overestimate rabies deaths in places with high reporting of bites.

In 2018, Gavi added PEP to their vaccine investment portfolio, and soon countries will be able to apply for funding to expand access to PEP. Expanding access strategically by identifying places with limited access...

In Madagascar, Institut Pasteur de Madagascar (IPM) provides PEP at no-cost to patients at 31 clinics across the country. PEP is not available at any other public clinics or through the private sector. In addition, there is limited control in dog populations and the disease is endemic throughout the country. Due to the spatially restricted nature of PEP, geographic access is likely to be a major driver of disease burden within the country. To get spatial estimates of disease burden in this context, we assume that bite incidence reflects access to PEP rather than than differences in rabies incidence, using travel times to clinics as a predictor of reported bites. Then using a range of rabies incidence given endemic transmission with no mass dog vaccination (*GAVI paper*), we generate sub-national estimates of rabies burden in an adapted decision tree framework. Finally, using this same model pipeline, we explore the impacts of geographically expanding access to PEP in Madagascar on reducing the burden of human rabies deaths.

@@

### Bite patient data
We used a database of bite patient forms submitted to IPM from ARMC across the country between 2014 - 2017. These were individual patient data forms that were submitted as frequently as monthly to annually by clinics. Of the 31 existing ARMC, `r length(no_data) - 2` submitted fewer than 10 forms over the four years. Two clinics, the IPM ARMC and the Fianarantsoa ARMC had separate databases which were not available at the time of analysis. Overall, we had data from *`r 31 - length(no_data)`* clinics across the country (Fig 1A). These data include details of the administrative district of the bite patient's address and the date of reporting. We also had *27* months of data from the Moramanga District that were resolved to the commune level (the administrative level below the district).

@@

Previous work in the Moramanga District showed that low risk contacts with probable or confirmed rabies cases make up approximately 20% of patients reporting to ARMC (Rajeev et al. 2018). Generally, contacts present as clustered cases, so we excluded patients reporting on any dates that had greater than 3 standard deviations above the mean number of patients reporting per day (Figure 3A). We validated this method using the Moramanga ARMC data for which we had details on the type of exposure, and found that setting the threshold to 3 standard deviations (SDs) resulted in approximately 50% of known contacts excluded, with only 2% of non-contacts excluded (Figure S4). For the national data for which a subset of patient forms were explicitly noted to be contacts, we found that our exclusion criteria of 3 SDs identified *r round((ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 & exclude_bydate == 1]))/length(ctar_check %$% exclude_bycomment[exclude_bycomment == 1])**100, 2)"* of known contacts. We further excluded these known contacts which were not identified based on the daily distribution of patients, resulting in the exclusion of approximately *r round((ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 | exclude_bydate == 1]))/nrow(ctar_check)**100, 2)*% of patient records from the national data. We also compared this method to assuming that 20% of total bites at the district level were contacts.

@@

After excluding contacts and correcting for undersubmission of forms, our final dataset consisted of estimates of average bite incidence for `r nrow(exps_dist)` districts from *r length(unique(exps_dist$dist_catch))* catchments (Figure 4).

@@

```{r models ttimes, cache = TRUE}
## Exclude for comm data!
comm_covars <- filter(comm_data, exclude == 0)

### Running estimation
mada_mods_covars <- run.mods(bites = exps_dist$bites, names_bites = exps_dist$district,
                             pop_predict = c("flatPop"),
                             pop_vars = list(pop_dist = exps_dist$pop, pop_commune = comm_covars$pop),
                             covars = list(ttimes_district = exps_dist$ttimes_weighted_dist/60,
                                           ttimes_commune = comm_covars$ttimes_weighted/60), 
                             sum_it = c(FALSE, TRUE, FALSE, TRUE), run_type = "optim", label = "Mada")
mora_mods_covars <- run.mods(bites = exps_mora$bites, names_bites = exps_mora$commune,
                             pop_predict = c("flatPop"),
                             pop_vars = list(pop_dist = exps_mora$pop),
                             covars = list(ttimes_mora = exps_mora$ttimes_weighted/60), 
                             sum_it = c(FALSE, FALSE), run_type = "optim", label = "Mora")
mada_mods <- c(mada_mods_covars, mora_mods_covars)

## profile CIs
prof_CI <- lapply(mada_mods, confint)
bite_mods <- as.data.frame(list(beta = mapply(function(x) coef(x)["beta"], mada_mods),
                                intercept = mapply(function(x) coef(x)["intercept"], mada_mods),
                                beta_upper = mapply(function(x) x["beta", 2], prof_CI),
                                beta_lower = mapply(function(x) x["beta", 1], prof_CI),
                                intercept_lower = mapply(function(x) x["intercept", 1], prof_CI),
                                intercept_upper = mapply(function(x) x["intercept", 2], prof_CI),
                                likelihood = mapply(function(x) logLik(x), mada_mods),
                                model = mapply(function(x) attributes(x)$label, mada_mods)))
## Predictions: fixed
ttimes_plot <- seq(0, 25, by = 0.5)
distance_plot <- seq(0, 200, by = 2.5)
pop_plot <- seq(1000, 1e6, length.out = 30)

library(iterators)
preds_fixed <- run.predicts.fixed(covar_df = bite_mods, pop_fixed = 1e5, pop_seq = pop_plot)
preds_data <- run.predicts.data(covar_df = bite_mods, run_type = "predict.data")
preds_raw <- run.predicts.data(covar_df = bite_mods, run_type = "predict.raw")

ggplot(preds_fixed, aes(x = covar, y = predicted, color = scale)) + 
  geom_line() + 
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = scale), color = NA, alpha = 0.5) +
  scale_color_manual(values = c("navy", "blue", "purple")) +
  scale_fill_manual(values = c("blue", "lightblue", "pink")) +
  xlab("Travel time (hrs)") +
  ylab("Reported bites per \n 100,000 persons")

ggplot(preds_data, aes(x = observed, y = predicted, color = scale)) + 
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "grey") +
  expand_limits(y = c(0, 4000)) +
  scale_color_manual(values = c("navy", "blue", "purple"))

@@

@@
